VOCC outline

### 2020-04-23 ms/ms.Rmd
more results and figure edits

# Introduction

@@

Fishery-independent survey data is often collected to aid in stock assessments and fisheries management.
In Canadian Pacific waters, randomized bottom trawl surveys outfitted with dissolved oxygen and temperature collecting CTD sensors, have been run for about a decade now.
During this time bottom temperatures have tended to rise everywhere, while the direction of change in the availability of oxygen has been spatially variable.
In addition to estimates of biomass density for each species captured, data on size distributions and reproductive maturity are also collected for most species.
***We use these concurrent estimates of both climate and biomass to assess the extent to which groundfish densities are changing in response to climate at the local scales most relevant to management***.
In order to better understand the extent to which the temporal and spatial scales considered are appropriate for different species within this fish community, we also assess the concordance between the life histories and apparent sensitivities across species and maturity classes.
Given the relatively short timescale encompassed in this survey, we expect *stronger patterns in shorter-lived, more pelagic species, and immatures.*

<!-- TODO: Discuss with sean & chris rooper -->

# Methods

Example Figure reference: Fig. \@ref(fig:climate-maps).

## Survey data

@@

## Estimate spatiotemporal variation in adult and immature biomass densities

@@

TODO: DESCRIBE MODEL VALIDATION FOR EACH SPECIES AND CLIMATE MODEL

## Estimate climate velocities

@@

<!-- TODO: Explain calculation of gradient velocity-->
<!-- TODO: Explain collapsing to max of 100 km ... and update code and models to do this more precisely-->

@@

**Link biomass change with climate**

To explore the potential for change in climate to predict local percent changes in biomass density, we used similar spatial GLMMs in order to control for spatial autocorrelation amongst cells. For each maturity class of each species the model estimates an independent random spatial field was and with and without either genus or family as possible phylogenetic grouping factors. .
For each species, only the fewest cells that encompass 95% of total mean biomass were included.
In all models mean biomass density was included as a covariate in order to reduce the influence of changes occurring only at either the highest or lowest densities for a particular species.
We attempt to predict change in biomass of each species as a function of local climate change (just temperature, just DO, or both temperature and DO) and an interaction between the mean climate of each cell and its local change estimate.
We model both biomass and climate as either velocities or as raw trends<!-- TODO: should (Fig. \@ref(fig:climate-maps)) be here?-->. 
In addition we attempt to assess the importance of the gradient component in velocity estimates by adding gradient to the model based on trends.
All covariates were scaled to their standard deviations.
Covariates capturing cell means were also centered <!-- If fishing included: (except fishing effort---so that estimates can be interpreted as in the absence of fishing)-->, but change variables were kept uncentred because an intercept of 0 change is more intuitive to interpret.
Models exploring change as simple trends use Gaussian distributions, while velocities are heavy-tailed and therefore modeled using a student-t (degrees of freedom fixed at 7). 
Trend models converged with 500 knots and both temperature and DO, while velocity models would only converge with one environment variable at a time and 200 knots. <!-- TODO: make sure knots are explained somewhere... -->
In addition to checking the spatial distribution of model residuals, we simulated fake biomass trends for each species (using variance and spatial correlation parameters unique to each species) and reran each model 3 times <!-- TODO: simulate null model 10 times? --> and compared the distribution of effect sizes and number of sig effects in our simulated null models with the actual estimates. 
These simulated biomass trends were divided by the real spatial gradients in biomass in order to simulate biotic velocities. <!-- TODO: where to report on NULL results? -->

In order to better understand the extent to which the temporal and spatial scales considered are appropriate for different species within this groundfish community, we also assess the concordance between the life histories and apparent sensitivities across species and maturity classes.

# Results

Our models suggest that from 2008 to 2018, survey season seafloor temperature increased across almost the entirety of Canada's Pacific shelf region (95% interquartile range of `r signif(quantile(alldata$temp_trend, c(0.025, 0.975))[[1]], digits = 2)` to `r signif(quantile(alldata$temp_trend, c(0.025, 0.975))[[2]], digits = 2)`), with warming tending to be most pronounced in shallower areas (Fig. \@ref(fig:climate-maps)<!-- TODO: should letters be added and included here? eg. "C" -->). 
In contrast, change in seafloor DO was much more variable in the direction of change (95% interquartile range of `r signif(quantile(alldata$DO_trend, c(0.025, 0.975))[[1]], digits = 2)` to `r signif(quantile(alldata$DO_trend, c(0.025, 0.975))[[2]], digits = 2)`), with both the greatest increases also occurring in the shallowest areas and the greatest decreases occuring across a wide range of depths <!-- TODO: should letters be added and included here? eg. (Fig. \@ref(fig:climate-maps)D) -->. 
When these local changes are placed in their geographic context in the form of a gradient-based velocity estimate, the majority of the highest velocities occur in the bottom of troughs in Queen Charlotte Sound and Dixon Entrance <!-- TODO: should letters be added and included here? eg. (Fig. \@ref(fig:climate-maps)E & F) -->.

@@

Across species, the most common patterns were that 1) the impact of warming on biomass change was more negative (or at least less positive) in locations with higher mean temperatures than in cooler locations (e.g. Widow Rockfish, mature Shortspine Thornyhead, and mature English and Flathead Sole), 2) the impact of increasing DO levels was more negative in locations with lower mean DO levels than in locations with higher mean DO levels (e.g. Dogfish, mature Sablefish, Widow Rockfish, and English and Flathead Sole), and 3) overall immature abundances were more likely than mature abundances to show negative relationships (eg. Sablefish, Widow Rockfish, and Dover Sole (Fig. \@ref(fig:trend-chopsticks)). 
<!-- TODO: checking this with model including age as fixed effect (mature biomass as the intercept)... --> 
There were no strong patterns of similarity amoungst species belong to the same genus or family (
TODO: should we mention here that they were included as random effects in trial models, but not the final model?
) as illustrated by relationships with biotic trends that were negative at high temperatures (Canary Rockfish and English Sole) or positive at low temperatures (eg. Widow Rockfish and Flathead Sole) occuring in members of both the largest families represented in our analysis, Sebastidae (rockfish) and Pleuronectidae (righteye flounders; Fig. Fig. \@ref(fig:worm-plots and \@ref(fig:species-map-chop-panel)). 
Likewise, most species in all families showed relatively little impact of DO change when DO levels were high, but a somewhat counter-intuitive decline in biomass if DO increased when mean DO was at it's lowest (\@ref(fig:trend-chopsticks). 
However, several species do show increasing biomass with increasing DO across their distributions and most of these are rockfish (mature Bocaccio, Greenstripe, Redstripe, Yelloweye, Yellowmouth and Yellowtail; immature Sharpchin Rockfish; and the one exception, immature Sablefish).
The few exceptions to these pattern include a strong positive effect of increasing DO at low mean levels for adult Splitnose Rockfish, and both maturity classes of Shortspine Thornyheads, all populations occupying some of the deeper habitats. 
However, three species averaging even deeper in their mean occurance show the more common pattern (mature Sablefish and both maturity classes of the Rougheye/Blackspotted Rockfish complex), or no relationship at all with DO change (Shortraker Rockfish).

<!-- TODO: Figure out what to say about velocity "NULLs", if anything...
This spatial pattern largely remains for simulated biotic velocities, which results in models of simulated biotic velocity that have a similar number of significant effects as the model based on real biomass changes. 
These models often tend to show more significant effects for flatfish species, possibly due to distributions that encompass a greater proportion of flatter spatial gradients for both climate and biomass (TODO see Supplementary...). 
Both climate variables and species distributions were modeled in response to seafloor depth ...
-->
<!-- TODO:  Species of interest...
redbanded - pretty normal response to temp and DO (sig -ve for adults only) 
        widow rockfish  - strong interactions
canary - not interaction, -ve temp none for DO
shortspine thornyhead - both temp (increasing good) and DO (increasing good)
dogfish - increasing DO and warming both bad
pacific cod - 
English - warming neutral when cool, bad when warm
Dover - warming good when it's cool, neutral when warm  
(arrowtooth shows similar pattern but likely influenced strongly by fishing... could include for that reason though?)
sablefish - warming good, increasing DO not
-->

# Discussion

# Acknowledgements

# References

@@

# Figures

@@

```{r coef-scatterplots, fig.cap="Main effect coefficients for relationship between biotic trends and changes in climate for species with different life history strategies. Mean age and depth are calculated across all sampled individuals of each maturity class (as defined by the length at 50 percent maturity threshold) within a species. Growth rate is estimated as the length at maturity for females divide by the 0.95 quantile of ages recorded for immatures of each species.", out.width="8in"}
include_graphics(here("ms", "figs", "coef-scatterplots.pdf"))
```

@@

# Supporting Information {-}

@@

# Supporting Methods

@@

# Supporting Figures and Tables

Supplemental figures:
(not yet done)
-   flow diagram
-   spatial residuals
-   null model stuff
-   pairs plot of predictors
-   plot vocc regression residuals against the main predictors... (and not in model and fitted vs. residuals plot; but don't necessarily include in paper)

@@

@@

```{r supp-slope-by-depth, fig.cap="Shape of relationship between estimated mean dissolved oxygen and mean depth of each 4x4 km grid cell (pattern is vary similar for temperature so not illustrated) for comparison with slopes for the environmental extremes experienced at the deeper depths showing expected tendancy towards increasing variation in strength of responses amongst species with deeper mean depth of occurrence. Outliers amongst immatures are Widow Rockfish (both DO and temperature) and Bocaccio (temperature).", out.width="4in"}
include_graphics(here("ms", "figs", "supp-slope-by-depth.png"))
```

@@

>>>>

# **Introduction**

@@

Fishery-independent survey data is often collected to aid in stock assessments and fisheries management.
In Canadian Pacific waters, randomized bottom trawl surveys outfitted with dissolved oxygen and temperature collecting CTD sensors, have been run for about a decade now.
During this time bottom temperatures have tended to rise everywhere, while the direction of change in the availability of oxygen has been spatially variable.
In addition to estimates of biomass density for each species captured, data on size distributions and reproductive maturity are also collected for most species.
***We use these concurrent estimates of both climate and biomass to assess the extent to which groundfish densities are changing in response to climate at the local scales most relevant to management***.
In order to better understand the extent to which the temporal and spatial scales considered are appropriate for different species within this fish community, we also assess the concordance between the life histories and climate sensitivities across species and maturity classes.
Given the relatively short timescale encompassed in this survey, we expect *stronger patterns in shorter-lived, more pelagic species (TODO: how do I assess this?), and immatures.*

<!-- TODO: need an index of pelagic-ness? 
Need to add mention of growth rates?
Do I need why?
Discuss further with sean & chris rooper -->

# **Methods**

## *Survey data*

@@

## *Estimate spatiotemporal variation in adult and immature biomass densities*

@@

<!-- TODO: DESCRIBE MODEL VALIDATION FOR EACH SPECIES AND CLIMATE MODEL -->

## *Estimate climate velocities*

@@

<!-- TODO: Explain calculation of gradient velocity -->
<!-- TODO: Explain collapsing to max of 100 km ... and update code and models to do this more precisely-->

@@

## *Link biomass change with climate*

To explore the potential for change in climate to predict local percent changes in biomass density, we used similar spatial GLMMs in order to control for spatial autocorrelation amongst cells. For each maturity class of each species the model estimates an independent random spatial field was and with and without either genus or family as possible phylogenetic grouping factors. .
For each species, only the fewest cells that encompass 95% of total mean biomass were included.
In all models mean biomass density was included as a covariate in order to reduce the influence of changes occurring only at either the highest or lowest densities for a particular species.
We attempt to predict change in biomass of each species as a function of local climate change (just temperature, just DO, or both temperature and DO) and an interaction between the mean climate of each cell and its local change estimate.
We model both biomass and climate as either velocities or as raw trends <!-- TODO: should (Fig. \@ref(fig:climate-maps)) be here?-->. 
In addition we attempt to assess the importance of the gradient component in velocity estimates by adding gradient to the model based on trends.
All covariates were scaled to their standard deviations.
Covariates capturing cell means were also centered <!-- If fishing included: (except fishing effort---so that estimates can be interpreted as in the absence of fishing)-->, but change variables were kept uncentred because an intercept of 0 change is more intuitive to interpret.
Models exploring change as simple trends use Gaussian distributions, while velocities are heavy-tailed and therefore modeled using a student-t (degrees of freedom fixed at 7). 
Trend models converged with 500 knots and both temperature and DO, while velocity models would only converge with one environment variable at a time and 200 knots. <!-- TODO: make sure knots are explained somewhere... -->
In addition to checking the spatial distribution of model residuals, we simulated fake biomass trends for each species (using variance and spatial correlation parameters unique to each species) and reran each model 3 times <!-- TODO: simulate null model 10 times? --> and compared the distribution of effect sizes and number of sig effects in our simulated null models with the actual estimates. 
These simulated biomass trends were divided by the real spatial gradients in biomass in order to simulate biotic velocities. <!-- TODO: where to report on NULL results? -->

In order to better understand the extent to which the temporal and spatial scales considered are appropriate for different species within this groundfish community, we also assess the concordance between the life histories and climate sensitivities across species and maturity classes.
Life history parameters were estimated from all individual fish measured in these same trawl surveys. 
For all immatures and adults sampled for each species, we calculated the average depth at which they were captured, the average age estimated from otoliths or fin clips, and an approximation of immature female growth rates as the length at maturity for females divided by the 0.95 quantile of the ages of immature females. 
A similar value was calculated for immature males, but it only differed substantially for a couple of the fastest growing species and did not change any overall patterns.

# **Results**

Our models suggest that from 2008 to 2018, survey season seafloor temperature increased across almost the entirety of Canada's Pacific shelf region (95% interquartile range of `r signif(quantile(alldata$temp_trend, c(0.025, 0.975))[[1]], digits = 2)` to `r signif(quantile(alldata$temp_trend, c(0.025, 0.975))[[2]], digits = 2)`), with warming tending to be most pronounced in shallower areas (Fig. \@ref(fig:climate-maps) <!-- TODO: should letters be added and included here? eg. "C" -->). 
In contrast, change in seafloor DO was much more variable in the direction of change (95% interquartile range of `r signif(quantile(alldata$DO_trend, c(0.025, 0.975))[[1]], digits = 2)` to `r signif(quantile(alldata$DO_trend, c(0.025, 0.975))[[2]], digits = 2)`), with both the greatest increases also occurring in the shallowest areas and the greatest decreases occuring across a wide range of depths <!-- TODO: should letters be added and included here? eg. (Fig. \@ref(fig:climate-maps)D) -->. 
When these local changes are placed in their geographic context in the form of a gradient-based velocity estimate, the majority of the highest velocities occur in the bottom of troughs in Queen Charlotte Sound and Dixon Entrance <!-- TODO: should letters be added and included here? eg. (Fig. \@ref(fig:climate-maps)E & F) -->.

@@

<!-- TODO: Figure out what to say about velocity "NULLs", if anything...
This spatial pattern largely remains for simulated biotic velocities, which results in models of simulated biotic velocity that have a similar number of significant effects as the model based on real biomass changes. 
These models often tend to show more significant effects for flatfish species, possibly due to distributions that encompass a greater proportion of flatter spatial gradients for both climate and biomass (TODO see Supplementary...). 
Both climate variables and species distributions were modeled in response to seafloor depth ...
-->

Across species, the most common patterns were that 1) the impact of warming on biomass change was more negative (or at least less positive) in locations with higher mean temperatures than in cooler locations (e.g. Widow Rockfish, mature Shortspine Thornyhead, and mature English and Flathead Sole), 2) the impact of increasing DO levels was more negative in locations with lower mean DO levels than in locations with higher mean DO levels (e.g. Dogfish, mature Sablefish, Widow Rockfish, and English and Flathead Sole), and 3) overall immature abundances were more likely than mature abundances to show negative relationships (eg. Sablefish, Widow Rockfish, and Dover Sole (Fig. \@ref(fig:trend-chopsticks)). 
<!-- TODO: checking this with model including age as fixed effect (mature biomass as the intercept)... --> 
There were no strong patterns of similarity amoungst species belong to the same genus or family (TODO: should we mention here that they were included as random effects in trial models, but not the final model?) as illustrated by relationships with biotic trends that were negative at high temperatures (Canary Rockfish and English Sole) or positive at low temperatures (eg. Widow Rockfish and Flathead Sole) occuring in members of both the largest families represented in our analysis, Sebastidae (rockfish) and Pleuronectidae (righteye flounders; Fig. Fig. \@ref(fig:worm-plots and \@ref(fig:species-map-chop-panel)). 
Likewise, most species in all families showed relatively little impact of DO change when DO levels were high, but a potentially counter-intuitive decline in biomass if DO increased when mean DO was at it's lowest (\@ref(fig:trend-chopsticks). 
However, several species, mostly rockfish, do show increasing biomass with increasing DO across their distributions (mature Bocaccio, Greenstripe, Redstripe, Yelloweye, Yellowmouth and Yellowtail Rockfishes, and immature Sharpchin Rockfish and Sablefish).
The few exceptions to these pattern include a strong positive effect of increasing DO at low mean levels for adult Splitnose Rockfish, and both maturity classes of Shortspine Thornyheads, all groups occupying some of the deeper habitats (Fig. \@ref(fig:slope-by-depth)). 
However, three species with deeper mean occurence show the more common pattern (mature Sablefish and all of the Rougheye/Blackspotted Rockfish complex), or no relationship at all with DO change (Shortraker Rockfish).

Overall climate sensitivities did correlate with life history traits and were strongest for the youngest populations (Fig. \@ref(fig:coef-scatterplots)). 
This included overall more negative responses to both warming and increasing DO in populations with mean age less than ten years (Fig. \@ref(fig:coef-scatterplots) TODO: add panel letters?). 
Consistent with this, the strongest negative effects of warming (estimated for the warmest parts of a species distribution) were in species occurring at intermediate depths, while most species with mean encounter depths beyond 290 m appeared to increase in abundance with greater warming (Fig. \@ref(fig:slope-by-depth) TODO: add panel letter A). 
In contrast, both strong negative and positive responses to changes in DO occur across a range of depth preferences, although increasing DO in the lowest DO locations is mostly negative for species averaging shallower than 150 m (Fig. \@ref(fig:slope-by-depth) TODO: add panel letter B).   
Finally, species with faster growth rates in immature females (cm/year) tended to have more negative responses to increases in temperature or DO (Fig. \@ref(fig:coef-scatterplots) TODO: add panel letters?).

<!-- TODO:  Species of interest...
redbanded - pretty normal response to temp and DO (sig -ve for adults only) 
        widow rockfish  - strong interactions
canary - not interaction, -ve temp none for DO
shortspine thornyhead - both temp (increasing good) and DO (increasing good)
dogfish - increasing DO and warming both bad
pacific cod - 
English - warming neutral when cool, bad when warm
Dover - warming good when it's cool, neutral when warm  
(arrowtooth shows similar pattern but likely influenced strongly by fishing... could include for that reason though?)
sablefish - warming good, increasing DO not
-->

# **Discussion**

# **Acknowledgements**

# **References**

@@

# **Figures**

@@

```{r slope-by-depth, fig.cap="Slopes of the relationship between changes in biomass and climate for the environmental extremes expected to have the greatest physiological impact (highest temperatures [A] and lowest DO [B]) plotted against each species mean depth of occurrence. Smoothed (solid lines represent GAM) relationship between estimated temperature (right axis and blue) and DO (left axis and purple) with mean depth in each 4x4 km grid cell (dots). An increase in variablity of slopes for DO change occurs for species whose mean depth is at or below depths where estimated mean DO levels approach hypoxic conditions (dashed line at <1.4 ml/L).", out.width="4in"}
include_graphics(here("ms", "figs", "slope-by-depth.png"))
```

```{r coef-scatterplots, fig.cap="Main effect coefficients for relationship between biotic trends and changes in climate for species with different life history strategies. Mean age and depth are calculated across all sampled individuals of each maturity class (as defined by the length at 50 percent maturity threshold) within a species. Growth rate is estimated as the length at maturity for females divide by the 0.95 quantile of all otolith or fin-ray ages from immature females of each species. Separate regression slopes are plotted for each maturity class except in relation to mean age given that maturity is largelly a function of age.", out.width="8in"}
include_graphics(here("ms", "figs", "coef-scatterplots.pdf"))
```

@@

# **Supporting Information** {-}

@@

# **Supporting Methods**

@@

# **Supporting Figures and Tables**

<!-- TODO: Possible additional figures:
-   flow diagram
-   null model stuff
-   pairs plot of predictors
-   plot vocc regression residuals against the main predictors... 
-   fitted vs. residuals plot
-->

@@

```{r global-coefs-w-age, fig.cap="Overall coefficient estimates for model with mature as the intercept and two-way interactions with immature.", out.width="3in"}
include_graphics(here("ms", "figs", "supp-global-coefs-w-age.pdf"))
```

@@

```{r mat-residuals, fig.cap="Spatial residuals of mature biomass of each species in the trend-based model of biomass change predicted by environmental change.", out.width="7in"}
include_graphics(here("ms", "figs", "supp-mat-residuals.pdf"))
```

@@

```{r imm-residuals, fig.cap="Spatial residuals of immature biomass in the trend-based model of biomass change predicted by environmental change.", out.width="7in"}
include_graphics(here("ms", "figs", "supp-imm-residuals.pdf"))
```
